---
output: html_document
---
# Reproducible Research: Peer Assessment 1


### 1. Load the data from course website; (i.e. read.csv())

```{r, echo=TRUE}
activity<- read.csv("activity.csv")
```


### 2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r, echo=TRUE}
activity$date <- as.Date(activity$date, format = "%Y-%m-%d")
activity$interval <- as.factor(activity$interval)
```


## What is mean total number of steps taken per day?


### 1.Calculate the total number of steps taken per day.

```{r, echo=TRUE}
steps_daily <- aggregate(steps ~ date, activity, sum)
colnames(steps_daily) <- c("date","steps")
head(steps_daily)
```


### 2. Make a histogram of the total number of steps taken each day.

```{r, echo=TRUE}
library(ggplot2)
ggplot(steps_daily, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1500) + 
        labs(title="Summary of Steps Taken per Day (Oct-Nov 2012)", 
        x = "Total Number of Steps per Day", y = "Daily Frequency (intervals)" ) +theme_bw()
```


### 3. Calculate and report the mean and median of the total number of steps taken per day

```{r, echo=TRUE}
mean_steps_daily   <- mean(steps_daily$steps, na.rm=TRUE)
median_steps_daily <- median(steps_daily$steps, na.rm=TRUE)
```

Print:
```{r, echo=TRUE}
mean_steps_daily
```
```{r, echo=TRUE}
median_steps_daily
```
The average (mean) steps taken per day is 10766.19 while the median is 10765.


## What is the average daily activity pattern?

Prior to plotting, convert the steps taken by the 5-minute intervals into integers 



```{r, echo=TRUE}
interval_steps <- aggregate(activity$steps, 
                  by = list(interval = activity$interval),
                  FUN=mean, na.rm=TRUE)

interval_steps$interval <- as.integer(levels(interval_steps$interval)[interval_steps$interval])
        as.integer(levels(interval_steps$interval)[interval_steps$interval])
colnames(interval_steps) <- c("interval", "steps")
```


### 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}
ggplot(interval_steps, aes(x=interval, y=steps)) +   
        geom_line(color="green", size=1.5) +  
        labs(title="Average Daily Activity Pattern (Oct-Nov 2012)", x="Interval (5-Minutes)", y="Average Number of steps") +  
        theme_bw()
```


### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
max_interval_steps <- interval_steps[which.max(interval_steps$steps),]
print(max_interval_steps)
```

The 835th interval has the most steps taken at 206 steps.
     


##Imputing Missing Values

To address possible bias caused by missing values (NA), identify the number of missing values and device a strategy to fill out these values. Create a new (modified) dataset that integrates the new values.


### 1. Total number of missing values:
Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo=TRUE}
missing_steps <- sum(is.na(activity$steps))
print(missing_steps)
```
There are a total of 2304 missing values.



### 2. Fill the missing values and create a new dataset with the proxy values

Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

```{r, echo=TRUE}
na_steps_proxy <- function(data, pervalue) {
        na_list <- which(is.na(data$steps))
        na_proxy <- unlist(lapply(na_list, FUN=function(idx){
                interval = data[idx,]$interval
                pervalue[pervalue$interval == interval,]$steps
        }))
        proxy_na_steps <- data$steps
        proxy_na_steps[na_list] <- na_proxy
        proxy_na_steps
}

activity_proxy <- data.frame(  
        steps = na_steps_proxy(activity, interval_steps),  
        date = activity$date,  
        interval = activity$interval)
```

Check if there still missing values in the new dataset: activity_proxy. 

```{r, echo=TRUE}
missing_steps_new <- sum(is.na(activity_proxy$steps))
print(missing_steps_new)

str(activity_proxy)
```


### 3. Plot a histogram of the total number of steps taken each day
Plot a historgram of the total number of steps taken per day with the proxy values 

```{r, echo=TRUE}
steps_daily_proxy<- aggregate(steps ~ date, activity_proxy, sum)
colnames(steps_daily_proxy) <- c("date","steps")

ggplot(steps_daily_proxy, aes(x = steps)) + 
       geom_histogram(fill = "green", binwidth = 1500) + 
        labs(title="Summary of Steps Taken per Day (Oct-Nov 2012)", 
             x = "Total Number of Steps per Day", y = "Daily Frequency (intervals)") + theme_bw() 
```

Calculate the new mean and median of the data set with proxy values

```{r, echo=TRUE}
mean_steps_daily_proxy   <- mean(steps_daily_proxy$steps, na.rm=TRUE)
median_steps_daily_proxy <- median(steps_daily_proxy$steps, na.rm=TRUE)
print(mean_steps_daily_proxy)
print (median_steps_daily_proxy)
```

Note that the new mean and median are now equal at 10766.189; compared to the previous dataset with missing values. However, it is important to note that the value of the mean has not changed thus replacing the missing values has no impact on the estimates of computing the total daily number of steps.



##Are there differences in activity patterns between weekdays and weekends?



### 1.) Using the new dataset, add a new column indicating what type of day the observations were taken (Weekday or weekend)

```{r, echo=TRUE}
activity_proxy$dateType <-  ifelse(as.POSIXlt(activity_proxy$date)$wday %in% c(0,6), 'weekend', 'weekday')

head(activity_proxy)
```


### 2.) Plot a graph that would show if there is any difference in the activity by type of day.

```{r, echo=TRUE}
activitybydaytype<- aggregate(steps ~ interval + dateType, activity_proxy, mean)

library(lattice)
xyplot(steps ~ interval|dateType, data = activitybydaytype,
       type = "l", layout = c(1,2),
       grid = TRUE,
              xlab="5-Minute Time Interval", ylab = "Average Number of Steps Taken",
       main= "Average Steps Taken in Weekdays vs Weekends")
```


The graph shows that activity during weekdays has the highest peak but weekends have more frequent peaks.

